DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2023-05-12 manuscript/manuscript.Rmd
add reviwers to acknowledgements
change figure and table refs for actual numbers

DateLife's core software applications are implemented in the R package `datelife`,
<!-- Its current stable version -- v`r packageVersion("datelife")`, is available for installation from the GitHub repository -->
<!-- The Comprehensive R Archive Network (CRAN) repository [@sanchez2022],  -->
and relies on functionalities from other biological R packages:
ape [@paradis2004],
bold [@Chamberlain2018],
geiger [@pennell2014geiger],
msa [@bodenhofer2015msa],
paleotree [@Bapst2012a],
phyloch [@Heibl2008],
phylocomr [@Ooms2018],
phytools [@Revell2012],
rotl [@Michonneau2016], and
taxize [@Chamberlain2013; @Chamberlain2018].
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.

@@

A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database. Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
These matching pruned chronograms are referred to as source chronograms.
Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

@@

R package `datelife` code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime_main}).
<!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
generating new trees: shows bladj speed;
Adding dates processing time: reflects more mrbayes speed than anything;
get_bold_otol_tree running time: reflects bold downloading data speed.-->
<!--Speed with different number of chronograms in database? NO-->
<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->

@@

We illustrate the DateLife workflow using a family within the passeriform birds encompassing the true finches, Fringillidae, as case study. On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Fig. \ref{fig:figure2}. As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy. The resulting summary chronogram is shown in Fig. \ref{fig:fringillidages}, and results from previous steps of the workflow are available as Supplementary Figures.
<!--We used `r cite_r("r-references.bib")` for all our analyses.-->

@@

**_Creating a search query.--_**
We chose 6 bird species within the Passeriformes. The sample includes
two species of cardinals:
the black-thighed grosbeak -- *Pheucticus tibialis* and
the crimson-collared grosbeak -- *Rhodothraupis celaeno*;
three species of buntings:
the yellowhammer -- *Emberiza citrinella*,
the pine bunting --  *Emberiza leucocephalos* and
the yellow-throated bunting --  *Emberiza elegans*;
and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.
Processing of input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy (OTT v3.3, June 1, 2021). For a detailed discussion on the state of the synonym, refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in figure \ref{fig:figure2}A, which would not have had any data otherwise.
<!--\textcolor{red}{EJM: I don't understand what this sentence means}
LLSR: it is related to the synonym found by tnrs processing, if we had used the synonym instead of the one accepted in the database, we would not have discovered that information
LLSR: I moved it here, it makes more sense here, I think!.
-->

@@

**_Searching the database.--_**
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms from 6 different studies (Fig. \ref{fig:figure2}B). Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).

@@

**_Summarizing search results.--_**
DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
As expected, more inclusive nodes (e.g., node "n1") have more variance in age data than less inclusive nodes (e.g., node "n5").
Summary age data for node "n2" were excluded as final calibration because they are older than age data of the more inclusive node, "n1" (Fig. \ref{fig:figure2}C4).

@@

**_Summarizing search results.--_**
The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
As explained in the "Description" section, a tree topology to summarize age data upon is required.
By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}). Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life’s synthetic tree v12.3 [@opentreeoflife2019synth].

@@

Age data from source chronograms was congruified to OpenTree's topology (Fig. \ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementary Table S2). For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig. \ref{fig:fringillidages}).

@@

<!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
We performed a cross validation test of a DateLife analysis using the Fringillidae source chronograms obtained above (Supplementary Figs. S2-S20).
We used as inputs for a DateLife analysis all individual tree topologies from each of the 19 source chronograms from 13 studies, treating their node ages as unknown.
We congruified node ages extracted from chronograms from all other studies upon the individual topologies, effectively excluding original ages from each topology. Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. \ref{fig:cvXY}).
For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips. Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
In general, topologically deeper nodes display the largest age variation between node ages from the original chronograms and ages summarized with DateLife.

@@

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
Alternative source chronograms available for the same taxa have potentially been estimated implementing different types of calibrations, which affects the resulting node age estimates.
For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study [@burns2014phylogenetics] were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs. \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs. S1, S5).
<!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->
Another source of conflict in estimated node ages can arise from different placements for the same calibration, which would imply fundamentally distinct evolutionary hypotheses [@antonelli2017supersmart].
For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group is known to significantly affect estimates of time of lineage divergence [@sauquet2013practical].
Finally, placement of calibrations also affects uncertainty of node age estimates. For example, nodes that are sandwiched between a calibrated node and a calibrated root have less freedom of movement and hence narrower confidence intervals [@vos2004reconstructing], which inflates precision for nodes without calibrations but does not necessarily improve accuracy of the estimated ages.

@@

The DateLife project was born as a prototype tool aiming to provide the services describe in this paper, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
We thank colleagues from the O'Meara Lab at the University
of Tennessee Knoxville for suggestions, discussions and software testing.
The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
that led to initial work on this project. The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
The Open Tree of Life project that provides the open, metadata rich repository of
trees used to construct DateLife's chronogram database.
The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.

>>>>

DateLife's core software applications are implemented in the R package `datelife`,
<!-- Its current stable version -- v`r packageVersion("datelife")`, is available for installation from the GitHub repository -->
<!-- The Comprehensive R Archive Network (CRAN) repository [@sanchez2022],  -->
and relies on functionalities from other biological R packages:
ape [@paradis2004],
bold [@Chamberlain2018],
geiger [@pennell2014geiger],
msa [@bodenhofer2015msa],
paleotree [@Bapst2012a],
phyloch [@Heibl2008],
phylocomr [@Ooms2018],
phytools [@Revell2012],
rotl [@Michonneau2016], and
taxize [@Chamberlain2013; @Chamberlain2018].
Figure 1 provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.

@@

A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database. Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
These matching pruned chronograms are referred to as source chronograms.
Total distance in units of million years (Myr) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Fig. 1).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

@@

R package `datelife` code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. 2).
<!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
generating new trees: shows bladj speed;
Adding dates processing time: reflects more mrbayes speed than anything;
get_bold_otol_tree running time: reflects bold downloading data speed.-->
<!--Speed with different number of chronograms in database? NO-->
<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->

@@

We illustrate the DateLife workflow using a family within the passeriform birds encompassing the true finches, Fringillidae, as case study. On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Figure 3. As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy. The resulting summary chronogram is shown in Figure 4, and results from previous steps of the workflow are available as Supplementary Figures.
<!--We used `r cite_r("r-references.bib")` for all our analyses.-->

@@

**_Creating a search query.--_**
We chose 6 bird species within the Passeriformes. The sample includes
two species of cardinals:
the black-thighed grosbeak -- *Pheucticus tibialis* and
the crimson-collared grosbeak -- *Rhodothraupis celaeno*;
three species of buntings:
the yellowhammer -- *Emberiza citrinella*,
the pine bunting --  *Emberiza leucocephalos* and
the yellow-throated bunting --  *Emberiza elegans*;
and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.
Processing of input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy (OTT v3.3, June 1, 2021). For a detailed discussion on the state of the synonym, refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in Figure 3A, which would not have had any data otherwise.
<!--\textcolor{red}{EJM: I don't understand what this sentence means}
LLSR: it is related to the synonym found by tnrs processing, if we had used the synonym instead of the one accepted in the database, we would not have discovered that information
LLSR: I moved it here, it makes more sense here, I think!.
-->

@@

**_Searching the database.--_**
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms from 6 different studies (Fig. 3B). Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table 1).

@@

**_Summarizing search results.--_**
DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3C), and congruified and mapped age data to nodes in this chosen topology (Table 1).
The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_; Fig. 3A) that would not have had any data otherwise due to name mismatch.
Age summary statistics per node were calculated (Table 2) and used as calibrations to date the tree topology using the BLADJ algorithm.
As expected, more inclusive nodes (e.g., node "n1") have more variance in age data than less inclusive nodes (e.g., node "n5").
Summary age data for node "n2" were excluded as final calibration because they are older than age data of the more inclusive node, "n1" (Fig. 3C4).

@@

**_Summarizing search results.--_**
The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
As explained in the "Description" section, a tree topology to summarize age data upon is required.
By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. 5). Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life’s synthetic tree v12.3 [@opentreeoflife2019synth].

@@

Age data from source chronograms was congruified to OpenTree's topology (Fig. 5B), reducing the age data set to 818 different data points (Supplementary Table S2). For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig. 4).

@@

<!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
We performed a cross validation test of a DateLife analysis using the Fringillidae source chronograms obtained above (Supplementary Figs. S2-S20).
We used as inputs for a DateLife analysis all individual tree topologies from each of the 19 source chronograms from 13 studies, treating their node ages as unknown.
We congruified node ages extracted from chronograms from all other studies upon the individual topologies, effectively excluding original ages from each topology. Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 5).
For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips. Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
In general, topologically deeper nodes display the largest age variation between node ages from the original chronograms and ages summarized with DateLife.

@@

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
Alternative source chronograms available for the same taxa have potentially been estimated implementing different types of calibrations, which affects the resulting node age estimates.
For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study [@burns2014phylogenetics] were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs. 4, 5; Supplementary Figs. S1, S5).
<!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->
Another source of conflict in estimated node ages can arise from different placements for the same calibration, which would imply fundamentally distinct evolutionary hypotheses [@antonelli2017supersmart].
For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group is known to significantly affect estimates of time of lineage divergence [@sauquet2013practical].
Finally, placement of calibrations also affects uncertainty of node age estimates. For example, nodes that are sandwiched between a calibrated node and a calibrated root have less freedom of movement and hence narrower confidence intervals [@vos2004reconstructing], which inflates precision for nodes without calibrations but does not necessarily improve accuracy of the estimated ages.

@@

We thank Isabel Sanmartín, Daniele Silvestro, Rutger Vos and an anonymous reviewer, for comments that greatly improved this manuscript.
The DateLife project was born as a prototype tool aiming to provide the services describe in this paper, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
We thank colleagues from the O'Meara Lab at the University
of Tennessee Knoxville for suggestions, discussions and software testing.
The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
that led to initial work on this project. The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
The Open Tree of Life project that provides the open, metadata rich repository of
trees used to construct DateLife's chronogram database.
The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
